---
title: "Lysates_metabar_for_PLOSone_LabProtocols"
output: html_document
date: "2022-11-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(repos = list(CRAN="http://cran.rstudio.com/"))
if (!require(devtools)) install.packages("devtools")
#devtools::install_github("yanlinlin82/ggvenn")
install.packages("wesanderson")
install.packages("UpSetR")
install.packages("vegan")
install.packages("reshape2") 
library('vegan')
library('UpSetR')
library('reshape2')
library('ggplot2')
library('ggvenn')
library("wesanderson")
```
This data come from the "15 samples" experiment

## Data formating
### Load raw data and remove 0's
```{r raw counts, include=FALSE}
tab <- read.delim("data/Metabar_counts_LYSATES.csv", sep=";", row.names = 1)
tab_no0 <- tab[rowSums(tab[])>0,]
write.table(tab_no0,"data/Metabar_counts_LYSATES_no0.tsv", sep="\t")
```

### Number of reads per sample: minimum, maximum, average:
```{r stats}
min(colSums(tab_no0))
max(colSums(tab_no0))
mean(colSums(tab_no0))
```

### Getting taxonomy file
Export ASV list to a separate file and get taxonomy assignement file for those ASVs
```{bash ASVs_new, include=FALSE}
cut -f1 data/Metabar_counts_LYSATES_no0.tsv > data/ASVs.txt
sed 's/"//g' data/ASVs.txt > ASV_list.txt
#grep -wFf ASV_list.txt /Users/elzbiwas/Desktop/August22_IBA_results/MBs_12.merged.taxa.tsv > Metabar_LYSATES_taxa.tsv
```

## Summary stats
### Total number of ASVs:
```{r ASVs}
length(rownames(tab_no0))
```

### Wet weight of a sample (proxy for biomass) and the number of reads per sample:
-- *Samples sorted by habitat and by sampling date (increasing)
-- *Wet-weight shown as white diamonds
```{r reads, echo=FALSE}
con <- read.delim("data/LYSATES_condition.csv", sep=";", row.names = 1)

p <- ggplot(data=con, aes(x=RunNumDate, y=TotalSeqs, fill=Habitat)) +
  geom_bar(stat="identity", color="black") + theme_minimal() +
  geom_point(data=con, aes(x=RunNumDate, y=Wet_weight*101000), shape=23, fill="white", color="black") +
  scale_fill_manual(values=wes_palette(n=3, name="GrandBudapest1")) +
  scale_x_continuous(name="Sample", breaks=con$RunNumDate,labels=con$Samples_Name) +
  scale_y_continuous(name="Number of reads", sec.axis = sec_axis(~./101000, name = "Wet weight [mg]")) +
  theme(axis.text.x = element_text(angle=90)) #+
#  ggtitle("Figure 2")
p
ggsave(filename = "figures/Fig2.jpg", device="jpeg", width = 9, height = 5.5, p)

```
## ASVs shared between habitats

```{r divide, echo=FALSE}
#Divide samples according to habitats
con <- read.delim("data/LYSATES_condition.csv", sep=";", row.names = 1)

W <- rownames(subset(con, con$Habitat=="Wetland"))
WetlandT <- subset(tab_no0[,W])
Wetland <- rownames(WetlandT[rowSums(WetlandT[])>0,])

G <- rownames(subset(con, con$Habitat=="Grassland"))
GrasslandT <- subset(tab_no0[,G])
Grassland <- rownames(GrasslandT[rowSums(GrasslandT[])>0,])

For <- rownames(subset(con, con$Habitat=="Forest"))
ForestT <- subset(tab_no0[,For])
Forest <- rownames(ForestT[rowSums(ForestT[])>0,])
```

### Venn diagram showing numbers of shared and habitat-unique ASVs
```{r venn, echo=FALSE}
x <- list(Forest=Forest, Grassland=Grassland, Wetland=Wetland)
v<- ggvenn(x, fill_color = wes_palette("GrandBudapest1"),
  stroke_size = 0.5, set_name_size = 6) #+
  #ggtitle("Figure 4")
v
ggsave(filename = "figures/Fig4.jpg", device="jpeg", width = 6, height = 5.5, v)
```


### Upset plot showing numbers of ASVs shared between different habitats
Not sure I will include it after all. 
I wanted to extend this plot with splitting each bar into fractions representative of different insect orders.
However, this has proven difficult to implement. Perhaps not worth the time.

## Ordination
### Calculating a non-metric multidimensional scaling (nMDS)
```{r NMDS_calculate, echo=FALSE, include=FALSE}
counts <- read.delim("data/Metabar_counts_LYSATES_no0.tsv", sep="\t", row.names = 1)
couch <- as.data.frame(t(counts))
relAbund <- decostand(couch, method = "total")
DistMatrix <- vegdist(relAbund, method = "bray")
Mat_rix <- as.matrix(DistMatrix, labels = T)
write.csv(Mat_rix, "data/B-C_distance_matrix.csv")

NMS <-  metaMDS(DistMatrix, distance = "bray", k = 3, trymax = 50, wascores = TRUE)
```

### Stress factor
```{r NMDS_calculate2, echo=FALSE}
NMS$stress
```
As a rule of thumb literature has identified the following cut-off values for stress-level:

Higher than 0.2 is poor (risks for false interpretation).
0.1 - 0.2 is fair (some distances can be misleading for interpretation).
0.05 - 0.1 is good (can be confident in inferences from plot).
Less than 0.05 is excellent (this can be rare).

### Plotting nMDS
```{r NMDS_Plot, echo=FALSE}
con2 <- read.delim("data/LYSATES_condition_reordered.csv", sep=";")
data.scores <- as.data.frame(scores(NMS))  
data.scores$sample <- con2$Sample_ID
data.scores$grp1 <- con2$Habitat
data.scores$grp2 <- con2$Date

n<- ggplot(data=data.scores, 
  aes(x=NMDS1, y=NMDS2, fill=factor(grp1), shape=factor(grp2), color=factor(grp1))) + 
  geom_point(size=3) +
  scale_shape_manual(values=c(21,22,23,24,25)) +
  scale_fill_manual(values = wes_palette("GrandBudapest1")) +
  scale_color_manual(values = wes_palette("GrandBudapest1")) +
  guides(fill = "none") +
  coord_equal() + labs(color="Habitat", shape="Sampling date") +
  theme_bw() #+
  #ggtitle("Figure 5")
n
ggsave(filename = "figures/Fig5.jpg", device="jpeg", width = 9, height = 5.5, n)
```

### ASVs grouped per order
```{r barplot-order, echo=FALSE}
orders <- read.delim("data/Counts_per_order_uncorrected.csv", sep=";", row.names = 1)
ord <- as.data.frame(t(orders))
ord$X <- rownames(ord)
lo <- melt(ord, id.vars = "X", variable.name = "Orders")

r<-ggplot(lo, aes(x = X, y=value, fill = Orders)) + 
  geom_bar(position="fill", stat = "identity", color="black") + 
  theme_bw() + theme(axis.text.x = element_text(angle = 0, hjust = 1)) +
  labs(y = "Fraction", x="") +
  #theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle=90)) #+
#  ggtitle("Fig 6 - Relative abundance of orders per sample")
r
ggsave(filename = "figures/Fig6.jpg", device="jpeg", width = 9, height = 5.5, r)
```

## Read abundance per sample 
### Number of reads per sample - uncorrected
```{r barplot2, echo=FALSE}
b<-ggplot(lo, aes(x = X, y=value, fill = Orders)) + 
  geom_bar(stat = "identity", color="black") + 
  #scale_fill_manual(values = palette) +
  theme_bw() + theme(axis.text.x = element_text(angle = 0, hjust = 1)) +
  geom_point(data=con, aes(x=RunNumDate, y=Wet_weight*100000), shape=23, fill="white", color="black") +
  scale_y_continuous(name="Number of reads", sec.axis = sec_axis(~./100000, name = "Wet weight [mg]")) +
  labs(y = "Number of reads", x="") +
  ggtitle("Uncorrected numbers of reads") +
  #theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle=90)) #+
#  ggtitle("Fig. 7 - ASVs grouped per order")
b
ggsave(filename = "figures/Fig7.jpg", device="jpeg", width = 9, height = 5.5, b)
```

### Number of reads per sample - corrected

```{r barplot4, echo=FALSE}
corr <- read.delim("data/Counts_Corrected_per_order.csv", sep=";", row.names = 1)
cor5 <- as.data.frame(t(corr))
cor5$X <- rownames(cor5)

lp <- melt(cor5, id.vars = "X", variable.name = "Orders")

spi<- ggplot(lp, aes(x = X, y=value, fill = Orders)) + 
  geom_bar(stat="identity", color="black") + 
  theme_bw() + theme(axis.text.x = element_text(angle = 0, hjust = 1)) +
  geom_point(data=con, aes(x=RunNumDate, y=Wet_weight*15), shape=23, fill="white", color="black") +
  scale_y_continuous(name="Number of reads", sec.axis = sec_axis(~./15, name = "Wet weight [mg]")) +
  labs(y = "Number of reads", x="") +
  #theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle=90)) +
  ggtitle("Spike-in corrected numbers of reads")
spi
ggsave(filename = "figures/Fig8.jpg", device="jpeg", width = 9, height = 5.5, spi)

```




